---
title: "Hispanic/Latino(a) Health Disparities in a US Sample Y2018"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
---

```{r setup, include=FALSE}
library(flexdashboard)
pacman::p_load(haven, flextable, data.table, ggthemes, sandwich, flexdashboard, ggplot2, ggExtra, grid, tidyverse, rio, here, modelsummary, janitor, ggThemeAssist, gghighlight, ggridges)
```

```{r}
mydata <- import(here("data/Latinx Health Behaviors and Health Outcomes dataset_7.14.2020.sav")) %>%  janitor::clean_names()

df0 <- mydata %>% 
  select(id, hisp_combine, hsi2_7_combine, hsi2total, hsi_health_stress, current_health_cond_dichot, current_health_cond_totalcount, amas_combine_comp_4_combine, amas_combine_speak_5_combine, currentobesitydichot, age_combine, lang_perf, black_combine, region_combine, household_living_combine, gen_status_self_combine, household_income_combine, gender_combine, health_insurance_combine) %>% 
  drop_na()
df_clean <- df0 %>% mutate(nationality = as.factor(hisp_combine), discrimination_level = as.factor(hsi2_7_combine), total_health_stress = as.numeric(hsi_health_stress), y_n_health_cond = as.factor(current_health_cond_dichot), eng_speak = as.factor(amas_combine_speak_5_combine),age_combine, lang_pref = as.factor(lang_perf), black_ident = as.factor(black_combine), region = as.factor(region_combine), household_size = as.factor(household_living_combine), generation_status = as.factor(gen_status_self_combine), household_income = as.numeric(household_income_combine), gender = as.factor(gender_combine), health_ins = as.factor(health_insurance_combine))
#str(df_clean)
#summary(df_clean$gender)
df_clean$gender <- factor(df_clean$gender,
                          levels= c(1,2,3,4),
                          labels = c("Female", "Male", "Trans/Other", "Other"))
levels(df_clean$gender)[levels(df_clean$gender)=="Other"] <-"Trans/Other"
#summary(df_clean$gender)
df_clean$health_ins <- factor(df_clean$health_ins,
                          levels= c(1,2,3),
                          labels = c("Uninsured", "Public Ins", "Private Ins"))
#summary(df_clean$health_ins)
df_clean$nationality <- factor(df_clean$nationality, 
                                   levels = c(2:7), 
                                   labels = c("Puerto Rican", "Dominican", "Mexican", "Cuban", "Other Hisp", "Other Spanish"))
#summary(df_clean$nationality)
levels(df_clean$nationality)[levels(df_clean$nationality)=="Other Spanish"] <-"Other Hisp"
df_clean$y_n_health_cond <- factor(df_clean$y_n_health_cond,
                                         levels = c(0,1),
                                   labels = c("No", "Yes"))
                                              
df_clean$generation_status <- factor(df_clean$generation_status,
                                         levels = c(1,2),
                                         labels = c("US Born","Foreign Born")
                                              )
```

Sidebar Title {.sidebar}
==========================

**Significance of Research**
Many institutions in the U.S. group individuals by language; the ethnic Hispanic/Latino(a) population is one such case. Aggregating based on the social construct of ethnicity ignores the health and cultural heterogeneity embedded in the Hispanic/Latino(a) population and has led to the exacerbation of health disparities (Nava et al., 2022). Identifying which groups bear a disproportionate burden of disease is the first step in better understanding individual and systemic disparities, such as inequities in access to care and social determinants of health. And with health disparities now exacerbated by COVID-19, there is a heightened need for research in this area. 

**Who**
In January 2018, data was collected from 798 Hispanic/Latino(a) adults (52% Mexican or Mexican American, 17% Puerto Rican, 8.5% Cuban, 3% Domincan, and 19% Other Hispanic/Latino such as Columbian, Venezuelan, Spanish) in the United States using the online survey platform, Qualtrics. Participants (N = 774) were age 18 years of age or older (M = 40, SD = 15.05). 

**Why** 
The purpose of the original study was to investigate potential contributors to health outcomes among Hispanic/Latino(a) adults in the United States.

In this Dashboard it is being used to answer the following research questions:

**RQ 1**
How does nativity and having a chronic health condition vary among nationality sub-groups of Hispanic/Latino(a) adults in the United States? 

**RQ 2**
How does health care access (i.e., health insurance status) and having a chronic health condition vary among sub-groups of Hispanic/Latino(a) adults in the United States? 

**About Data:**
Data used for this analysis is from previous research study conducted by Budd (2020) and can be found in the Harvard Dataverse.
[Dataset](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/NABLZX)

**Navigate:**

Page 1 - *Demographics*

Page 2 - *Health Disparity Visuals*

Page 3 - *References and Repository*

Page 1 {data-orientation=rows data-icon="fa-check"}
==============================

Row {data-height=50}
-----------------------------------------

Each of the following graphs show the demographics of this population sample ( _N_ = 774). We can see in **Chart B** that the majority of our sample is Mexican/Mexican American. 

Row {data-height=600}
-----------------------------------------

### Chart A 

```{r}
demographics <- ggplot(df_clean, aes(x=age_combine, group=nationality, fill=nationality)) +
    geom_density(adjust = 1.5, alpha=.4)+
  facet_wrap(~ gender)+
  labs(x = "Age 18-80 for Gender by Nationality", title = "Age Frequency by Gender and Nationality", y = "")+
  scale_color_colorblind()+
  theme_tufte()
demographics
```

Row {data-height=200}
-----------------------------------------------------------------------

This sample had 459 identifying females, 305 males, and 10 transgender or other gender identifying individuals. **Chart A** shows that in this sample transgender and other gender identifying folks were largely on the younger side (18- ~30 years) for Mexican individuals and young to middle aged (20-60 years) for Dominican individuals. 
**Charts A** and **B** show Mexican/Mexican American individuals are also predominately 20 to 60 years old with more younger males than females. Puerto Rican individuals are predominately between 40 and 60 years old, with a higher proportion of older females. Dominicans were also on the younger side (20 - 40 years old) compared to Puerto Rican, other Hispanic, and Cuban individuals. And there is a larger proportion of other Hispanic female identifying individuals between 55 and 65 years of age compared to males. In addition, the frequency of male and female individuals identifying as other Hispanic and Cuban, seem to mirror each other, however, Cubans report (on the next page) worse health than the other Hipanic identifying group. 

Row {data-height=400}
-----------------------------------------------------------------------

### Chart B

```{r}
demographics_2 <- ggplot(df_clean, aes(x=age_combine, group=nationality, fill=nationality)) +
    geom_density(adjust = 1.5, alpha=.3)+
  scale_color_colorblind()+
  labs( x= "Age 18-80", title= "Age Frequency by Nationality", 
        y = "")+
   scale_color_colorblind()+
  theme_tufte()
demographics_2

```

Row {data-height=400}
-----------------------------------------------------------------------
### Chart C

```{r}
ggplot(df_clean, aes(nationality)) +
  geom_bar(fill = "cornflowerblue")+
  labs(title = "Nationality Group Frequencies",
       y = "Count",
       x = "")+
  theme_tufte()

```

Row {data-height=50}
-----------------------------------------------------------------------

In **Chart C** we can really see the differences in the number of participants belonging to each group. With percentages of each population reported in the **Who** column to your left. This was my first attempt at representing demographics for this group, and **Charts A** and **B** added more to the demographics, such as age and gender.  

Page 2 {data-orientation=rows data-icon="fa-check"}
==============================

Row {data-height=500}
-----------------------------------------

### Chart A

```{r}
ggplot(df_clean, aes(y_n_health_cond, hsi_health_stress, color = nationality)) +
  geom_jitter()+
    labs( x = "Presence of Health Condition",
        y = "Healthcare Related Stress",
        color = "Nationality", title = "The more stress from healthcare discrimination", subtitle = "the more reports of having a chronic health condition") +
  scale_color_discrete(name = "Nationality",
                       breaks=c("Puerto Rican", "Dominican", "Mexican or Mexican America", "Cuban", "Other Hisp")) +
  theme_minimal()

```

Row {data-height=125}
-----------------------------------------

**Chart A** was the first graph I made looking at differences in the presence of a chronic health condition among diverse nationality subgroups of Hispanic/Latino(a) adults in this sample. I incorporated stress from discrimination experienced in the healthcare setting, to see the association between stress and the presence of a chronic health condition and how this varied for different Hispanic/Latino(a) subgroups. **Chart A** shows that Puerto Rican, Dominican, and Cuban individuals experience greater stress from health care discrimination and report having a chronic health condition more than other groups. 

Row {data-height=500}
-----------------------------------------

### Chart B

```{r}
ggplot(df_clean, aes(y_n_health_cond, generation_status, color = nationality)) +
  geom_jitter()+
    labs( x = "Presence of Health Condition",
        y = "Healthcare Related Stress",
        color = "Nationality", title = "US born Puerto Rican, Dominican, and Cuban individuals", subtitle = "report more chronic health conditions") +
  scale_color_discrete(name = "Nationality",
                       breaks=c("Puerto Rican", "Dominican", "Mexican or Mexican America", "Cuban", "Other Hisp")) +
  theme_minimal()
```

Row {data-height=75}
-----------------------------------------

**Chart B** is my first attempt at displaying differences in the presence of a chronic health condition for varying Hispanic/Latino subgroups and nativity (US born vs. foreign born). But it's hard to say with confidence which groups report more chronic health conditions than others with this graph. 

Row {data-height=500}
-----------------------------------------

### Chart C

```{r}
group_health_gen <- ggplot(df_clean) + 
  geom_bar(aes(y_n_health_cond, fill = nationality), position = "fill") + 
  facet_wrap(~ generation_status)+
  labs( x= "Y/N Presence of a Chronic Health Conditon by Nationality", 
        y= "Proportion", title = "US Born Report Having more Health Conditions than Foreign Born", fill= "Nationality")+
  scale_color_colorblind()+
  theme_tufte()
group_health_gen

```

Row {data-height=75}
-----------------------------------------

**Chart C** is getting closer to being able to identify differences more easily. The y-axis is also set to a proportion and this adds to the ease of understanding proportional differences, but a side-by-side visual would be the best. 

Row {data-height=500}
-----------------------------------------

### Chart D

```{r}
group_health_gen3 <- ggplot(df_clean) + 
  geom_bar(aes(y_n_health_cond, fill = generation_status), position = "dodge") + 
  facet_wrap(~ nationality)+
  labs( x= "Y/N Presence of a Chronic Health Conditon by Nationality", 
        y= "Count", title = "US Born Report Having more Health Conditions than Foreign Born", subtitle = "Except for Mexican/Mexican Americans and 'Other' Hispanic/Latino(a) individuals", fill = "Nativity")+
  scale_color_colorblind()+
  theme_tufte()
group_health_gen3
```

Row {data-height=275}
-----------------------------------------

**Charts C** and **D** above are my resulting graphs and show that Hispanic/Latino individuals, identifying as Cuban, Puerto Rican, and Dominican born in the United States, experience higher reports of having a chronic health condition than foreign born individuals belonging to the same group. In other words, nationality is associated with the presence of a chronic health condition for US born participants, with the likelihood of a reported chronic disease higher than expected for Puerto Rican, Domincan, and Cuban individuals and lower than expected for Mexicans/Mexican Americans and 'other' identifying Hispanic/Latino individuals in this sample. **Chart D** shows the same information as **Chart C**, but in a way that allows you to compare group differences in health and nativity more easily.

**Caution:**

Correlation does not equal causation. Being born in the United States doesn't cause some groups to have a chronic health condition more than others. There are many variables that have not been taken into consideration when evaluating the relationship between nationality, generation status, and health in this sample. Future research should look at other known issues related to negative health outcomes; such as income, age, lack of access to health care, and health care utilization. Referencing back to Page 1, Puerto Rican and Cuban individuals were largely older compared to some of the other groups (~40-80 years).


Row {data-height=500}
-----------------------------------------
### Chart E

```{r}
group_health_access <- ggplot(df_clean, aes(x = y_n_health_cond, fill = nationality)) + 
  geom_bar(positon = "dodge") + 
  facet_wrap(~ health_ins)+
  xlab("Y/N Presence of a Chronic Health Conditon by Insurance Status")+
  labs( title = "Health of Adult Hispanic/Latino Participants in the US by Insurance Status",
        y = "Count", fill = "Nationality")+
  scale_color_colorblind()+
  theme_tufte()
group_health_access

```

Row {data-height=135}
-----------------------------------------

**Chart E** shows differences in the presence of a chronic health condition by insurance status: Uninsured, Public Insurance, and Private Insurance. I added group nationality in the bars to help see health differences between groups, but the focus of this graph was to primarily visualize the differences in health outcomes by insurance. However, for varying Hispanic/Latino(a) groups and the differences in health outcomes, we can really see that Cuban and Puerto Rican individuals report having a chronic health condition more than the other groups _seemingly_ regardless of insurance status. I'd like to also note that for those identifying as 'other' Hispanic/Latino nationality, public health insurance _may_ be a protective factor in their health. Future research should investigate this difference more to help encourage policy makers to see the value in and potential (statistically speaking) protective nature of public health insurance among diverse communities. 

Page 3 {data-orientation=rows data-icon="fa-check"}
==============================

Row 
-----------------------------------------

### **Thanks!**
```{r}
# Create a vector of x and y values
x <- rnorm(1)
y <- rnorm(1)

# Create a plot with the x and y values
plot(x, y, xlab="", ylab="")

# Add the text "Thank You" at each data point
text(x, y, "Thank You!", col="#7E8FC7", cex=3.5)
```

Row 
-----------------------------------------

### **References**

Budd, E. L. (2020). Correlates of Health Behaviors and Outcomes among U.S. Latinx 
Adults. Harvard Dataverse, V1, UNF:6:I22oyR6qiaPV53LR37bjuw== 
[fileUNF]. https://doi.org/10.7910/DVN/NABLZX

Nava, A., Estrada, L., Gerchow, L., Scott, J., Thompson, R., & Squires, A. (2022). Grouping 
people by language exacerbates health inequities-The case of Latinx/Hispanic populations in the US. _Research in nursing & health_ , 45(2), 142–147. https://doi.org/10.1002/nur.22221

Williams, D. & Collins, C. (1995). US Socioeconomic and Racial Differences in Health: Patterns 
and Explanations. _Annu. Rev. Sociol_ , 21, 349-86. 

University of Oregon, EDLD 652 Winter 2023

**Repo**
[Github](https://github.com/bspinneruo/edld652_final.git)

